Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-02-13 ParisEM.tex
Overall revision

%Statistical inference for partially observed stochastic differential equations is a challenging task since some elementary quantities, such as transition probabilities, are not available explicitly.
Statistical inference for partially observed stochastic differential equations often requires to solve bayesian filtering and smoothing problems, i.e. the computation of the posterior distributions of sequences of hidden states given observations. Filtering refers to the estimation of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$, while smoothing stands for the estimation of the distribution of sequence of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p<\ell \le n$. 
These posterior distributions are crucial to compute maximum likelihood estimators of the unknown parameters defining the model using the observations $(Y_0,\ldots,Y_n)$ only.
% $\theta\mapsto \ell_{\theta}^{n}$ given by
%\begin{equation*}
%\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
%\end{equation*}  
%where $p_{\theta}$ is the complete data likelihood:
%\begin{equation*}
%p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.
%\end{equation*}
For instance, the E-step of the EM algorithm introduced in \cite{dempster:laird:rubin:1977} %iteratively builds a sequence of parameter estimates following two steps.
%following the two steps:
%\begin{enumerate}
%	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
%	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
%\end{enumerate}
%step 
boils down to the computation of a conditional expectation of an additive functional of the hidden states given all the observations up to time $n$. 
%\begin{multline*}
%Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
%+ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.
%\end{multline*}
Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states.
 See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.

The exact computation of these expectations is usually not possible in the case of partially observed diffusions. In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights. Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly. Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}. Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein. However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}. To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{fearnhead:papaspiliopoulos:roberts:2008}.  In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.\\
 The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. 
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. 
%In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. 
Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$. Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass. 
Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e. using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.

In this paper, instead of using fixed-lag techniques which introduce a systematic additional bias in the approximation of conditional expectations, we extend the use of  the PaRIS algorithm to partially observed SDE. 
The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles in the best cases, and at worse with quadratic complexity. 
The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by positive and bounded unbiased estimates. The proposed method results in a generalized random version of the PaRIS algorithm, hereafter named GRand PaRIS algorithm. The proposed algorithm apply to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation, belonging to one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
%We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied. To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}. These new GPE estimators use an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.

@@

\section{The generalized random PaRIS algorithm}
\label{sec:rwparis}
$(X_t)_{t\ge 0}$ is defined as a weak solution to the following SDE in $\mathbb{R}^d$:

@@

\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \rmd W_t\eqsp,

@@

where $(W_t)_{t\ge 0}$ is a standard Brownian motion. It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function. 
A popular choice of $A$  is $A = \log \pi/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$. See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi$. 
SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.\\
The solution of \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $\mathbb{R}^m$. 
For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,Y_k) = g_k(X_k)$. 
The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.\\ 
Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denote the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
\label{def:addfunc}
\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. 
The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,~Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.\\
At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. 
Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = \chi(\xi^{\ell}_0)g_0 (\xi^{\ell}_0)/\eta_0(\xi^{\ell}_0)$. 
For any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
\phi^N_{0}[h] = \frac{1}{\Omega_0^N} \sum_{\ell=1}^N \omega_0^{\ell} h \left(\xi^{\ell}_0 \right)\eqsp, \quad \Omega_0^N:= \sum_{\ell=1}^N \omega_0^{\ell}\eqsp.

>>>>

%Statistical inference for partially observed stochastic differential equations is a challenging task since some elementary quantities, such as transition probabilities, are not available explicitly.
Statistical inference for partially observed stochastic differential equations often requires to solve bayesian filtering and smoothing problems, i.e. the computation of the posterior distributions of sequences of hidden states given observations. Filtering refers to the estimation of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$, while smoothing stands for the estimation of the distribution of sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p<\ell \le n$. 
These posterior distributions are crucial to compute maximum likelihood estimators of the unknown parameters defining the model using the observations $(Y_0,\ldots,Y_n)$ only.
% $\theta\mapsto \ell_{\theta}^{n}$ given by
%\begin{equation*}
%\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
%\end{equation*}  
%where $p_{\theta}$ is the complete data likelihood:
%\begin{equation*}
%p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.
%\end{equation*}
For instance, the E-step of the EM algorithm introduced in \cite{dempster:laird:rubin:1977} %iteratively builds a sequence of parameter estimates following two steps.
%following the two steps:
%\begin{enumerate}
%	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
%	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
%\end{enumerate}
%step 
boils down to the computation of a conditional expectation of an additive functional of the hidden states given all the observations up to time $n$. 
%\begin{multline*}
%Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
%+ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.
%\end{multline*}
Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states.
 See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.

The exact computation of these expectations is usually not possible in the case of partially observed diffusions. In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights. Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly. Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}. Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein. However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}. To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{fearnhead:papaspiliopoulos:roberts:2008}.  In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.

 The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. 
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. 
%In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. 
Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$. Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass. 
Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e. using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.

In this paper, we extend the use of  PaRIS algorithm to partially observed SDE. 
The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles. The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates. The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights. The proposed generalized random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.
%We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied. To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}. These new GPE estimators use an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.

@@

\section{The generalized random PaRIS algorithm}
\label{sec:rwparis}
$(X_t)_{t\ge 0}$ is defined as a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion. It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function. 
A popular choice of $A$  is $A = \log \pi/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$. See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi$. 
SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein. The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $\mathbb{R}^m$. 
For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,Y_k) = g_k(X_k)$. 
The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.

Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
\[
\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
\]
For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denote the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
\begin{equation}
\label{def:addfunc}
\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
\end{equation}
when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. 
The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
\begin{equation}
\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,~Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
\end{equation}
The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.

At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. 
Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = \chi(\xi^{\ell}_0)g_0 (\xi^{\ell}_0)/\eta_0(\xi^{\ell}_0)$. 
For any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
\[
\phi^N_{0}[h] = \frac{1}{\Omega_0^N} \sum_{\ell=1}^N \omega_0^{\ell} h \left(\xi^{\ell}_0 \right)\eqsp, \quad \Omega_0^N:= \sum_{\ell=1}^N \omega_0^{\ell}\eqsp.
\]
Then, for $1\le k \le n$, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, the auxiliary particle filter of \cite{pitt:shephard:1999} samples pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles using an instrumental transition density $p_k$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$. Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computing following these steps:
\begin{enumerate}[-]
\item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_k(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
\item  associate the particle $\xi^{\ell}_k$ with the importance weight:
\begin{equation}
\label{eq:importance:weights}
\omega^{\ell}_k := \frac{\qk(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
\end{equation}
\end{enumerate} 
The expectation $\phi_{k}[h]$ is approximated by
\[
\phi^N_{k}[h] := \frac{1}{\Omega_t^N} \sum_{\ell=1}^N \omega_k^{\ell} h \left(\xi^{\ell}_k \right)\eqsp,\quad\Omega_k^N:= \sum_{\ell=1}^N \omega_k^{\ell}\eqsp.
\]
PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions. 
%The FFBS algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive. 
%An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.  
%Instead of computing new weights in the backward pass, the FFBSi  samples backward new trajectories from time $n$ to time $0$ among all the $N^{n+1}$ trajectories which can be chosen in $\{\xi^{\ell}_k\}$, $1\le \ell\le N$ and $0\le k\le n$. Contrary to the FFBS algorithm, the FFBSi algorithm samples trajectories and may be used to approximate joint smoothing distributions without increasing its complexity.
%Nevertheless, the computation of the transition probabilities to select particles in the backward pass makes it a $\mathcal{O}(N^2)$ algorithm. In the case where the transition probability $q_k$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}. 
It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. 
Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.

PaRIS algorithm relies on the following fundamental property of $T_k[H_k]$ when $H_k$ is as in \eqref{def:addfunc}:
\begin{align*}
T_k[H_k](X_k) &=\mathbb{E}\left[T_{k-1}[H_{k-1}](X_{k-1}) + h_{k-1}(X_{k-1},X_k)\middle|X_k,Y_{0:k-1} \right]\eqsp,\\
&= \frac{\int \phi_{k-1}(\rmd x_{k-1})q_{k-1}(x_{k-1},X_k)\left\{T_{k-1}[H_{k-1}](x_{k-1}) + h_{k-1}(x_{k-1},X_k)\right\}}{\int \phi_{k-1}(\rmd x_{k-1})q_{k-1}(x_{k-1},X_k)}\eqsp.
\end{align*}
Therefore, \cite{olsson:westerborn:2016} introduces sufficient statistics $\tau^i_k$ (starting with $\tau^i_0 = 0$, $1\le i\le N$), approximating $T_k[H_k](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$. First, replacing $\phi_{k-1}$ by $\phi^N_{k-1}$ in the last equation leads to the following approximation of $T_k[H_k](\xi^i_k)$:
\begin{equation}
\label{eq:Tk}
T_k^N[H_k](\xi_k^i) = \sum_{j=1}^N \Lambda_{k-1}^N(i,j)\left\{T_{k-1}[H_{k-1}](\xi_{k-1}^j) + h_{k-1}(\xi^j_{k-1},\xi^i_k)\right\}\eqsp, 
\end{equation}
where
\begin{equation}
\label{eq:Lambda}
\Lambda_{k}^N(i,\ell) = \frac{\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
\end{equation}
Then, instead of computing exactly these approximations which would lead to a complexity growing quadratically with $N$ because of the normalizing constant in \eqref{eq:Lambda}, PaRIS algorithm samples particles in the set $\{\xi^j_{k-1}\}_{j=1}^N$ with probabilities $\Lambda_{k}^N(i,\cdot)$ to approximate the expectation \eqref{eq:Tk} and produce $\tau^i_k$.
Choosing $\tilde{N}\ge 1$, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
\begin{enumerate}[(i)]
\item \label{it:PaRIS:filt} Run one step of a particle filter to produce $\{(\xi^{\ell}_k, \omega^{\ell}_k)\}$ for $1\le \ell \le N$.
\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_{k}^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k}^N(i,\cdot)$, given by \eqref{eq:Lambda}.
\item \label{it:PaRIS:smooth} Set
\[
\tau^{i}_{k+1} := \frac{1}{\widetilde{N}} \sum^{\widetilde{N}}_{\ell=1} \left\{ \tau^{J_{k}^{i,\ell}}_{k} + h_{k} \left(\xi^{J_{k}^{i,\ell}}_{k}, \xi^{i}_{k+1}\right)  \right\}\eqsp.
\]
\end{enumerate}
Then, \eqref{def:addfunc} is approximated by
\[
\phi_{0:n\vert n}^N = \frac{1}{\Omega_n^N}\sum_{i=1}^N \omega^{i}_n \tau_n^i\eqsp.
\] 
As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any fixed precision parameter $\tilde N$. However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$. As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the accept reject mechanism of \cite{douc:garivier:moulines:olsson:2011}.

In general situations, PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown. Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable. Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)&> 0~~\text{a.s}\eqsp,\\
\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)\middle| \mathcal{F}_{k+1}^N\right] &= \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
where, for all $0\le k \le n$, 
\mathcal{F}_{k}^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}),~1\le \ell\le N,~0\le u\le k\right\}\eqsp.
In the case where $\qk$ is unknown, the filtering weights in \eqref{eq:importance:weights} then become:

@@

\label{eq:random:weight}
\widehat{\omega}^{\ell}_k := \frac{\widehat{q}_k(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k;\zeta_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.

@@

Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM. If $\widehat{q}_k$ may be upper bounded almost surely by a constant $\hat{\sigma}_+$, this task is performed using the accept reject mechanism given in Algorithm~\ref{alg:AR:unknownq}. Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.\\
